Testing gene methylation enrichment analysis approaches using simulated data - summarise data

### 2023-11-03 figs/GSE158422_simulate_summary.Rmd
fig1

In this report we establish a new method for generating simulated data with known ground truth.
This will be used to test different gene methylation enrichment approaches systematically.

The general steps are:

1. Import GSE158422 data corresponding to control (non-tumour tissue).

2. From the 37 samples, create two groups of 18 samples.
One of these will be considered "control" and the other "case".

3. Create random gene sets that have similar sized to Reactome pathways.

3. Some gene sets will be selected to be differentially methylated.
Half of these will be hypermethylated and the others will be hypomethylated.

4. The changes will be incorporated into the "case" samples.

5. The enrichment analysis will be conducted.

6. The accuracy will be calculated.

@@

@@

# Geneset size = 100

```{r,gsa}

@@

@@

dev.off()

>>>>

Here is an explanation of the different objects:

0. GSA: gres

1. LAT: lacres

2. LTT: lactopres

3. LAW: nlacres

4. LAM: lacmres

5. ALT: alcres

6. ALW: nalcres

7. ALM: almres

@@

summarise <- function(x) {
  PREC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FP"]) )
  REC <- sum(x[,"TP"]) / ( sum(x[,"TP"]) + sum(x[,"FN"]) )
  F <- F1(PREC,REC)
  return(c("PRECISION"=PREC,"RECALL"=REC,"F1"=F))
}

@@

## Overall summary

Make a set of three barplots which summarise method performance over the range of group size and
delta values, for gene set sizes of 20, 50 and 200.

```{r,summary,fig.width=10,fig.height=4}

par(mfrow=c(1,3))

load("GSE158422_simulate020.Rdata")

l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x <- do.call(rbind,lapply(l,summarise))

barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="20 genes per set")
grid()

load("GSE158422_simulate050.Rdata")

l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x <- do.call(rbind,lapply(l,summarise))

par(mfrow=c(1,1))

barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="50 genes per set")
grid()

@@

l <- list("GSA"=gres2,"LAT"=lacres2,"LTT"=lactopres2,"LAW"=nlacres2,
  "LAM"=lacmres2,"ALT"=alcres2,"ALW"=nalcres2,"ALM"=almres2)

x <- do.call(rbind,lapply(l,summarise))

par(mfrow=c(1,1))

barplot(t(x),beside=TRUE,ylim=c(0,1.2),ylab="index", legend = colnames(x) ,main="100 genes per set")
grid()

## In depth analysis of GSA vs LAM

As GSA appears to be the best-performing existing method for enrichment analysis, it is a good idea
to compare it to LAM which is the best one out of the new methods evaluated here.
The following charts show the performance of GSA and LAM including precision and recall over the
range of group sizes and delta values when the sets contain 20, 50 and 100 genes each.

```{r,020_gsa_vs_lam}

# 20
load("GSE158422_simulate020.Rdata")

@@

# 50
load("GSE158422_simulate050.Rdata")

# 100
load("GSE158422_simulate100.Rdata")

par(mfrow=c(1,1))
